Subsidiary Recursion in Coq

### 2022-01-16 paper/paper.tex
complete draft
I will read over again, polish, may add more related work, then
circulate to the co-authors.

\subsection{Retractive-positive recursive types}
\label{sec:mu}

@@

\textbf{Termination.}
In some tools, like Coq, Agda, and Lean, termination is checked
statically, based on structural decrease at recursive calls.  Others,
like Isabelle/HOL, allow one to write recursions first, and prove
(possibly with automated help) their termination
afterwards~\cite{krauss}.

It has not escaped the notice of designers of ITPs that structural
recursion is not the only form of terminating recursion.  All the
mentioned tools provide support for well-founded recursion, where for
recursive calls, one must show that the parameter of recursion has
decreased in some well-founded order.

Subsidiary recursion can be seen as a generalization of \emph{nested
recursion}, which allows recursive calls of the form \verb|f (f x)|.
In subsidiary recursion, these are generalized to the form
\verb|f (g x)|, where \verb|g| could be \verb|f| or another
recursively defined function.

\textbf{Mendler encoding.}
Mendler introduced the basic idea of using universal abstraction to
support compositional termination checking; an accessible source
is~\cite{mendler91}.  He introduces a functor-generic recursor of type
\[
\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to X}
\]
\noindent We have adopted this idea to the constructor of the type
\verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
categorical perspective on Mendler-style recursion~\cite{uustalu99}.
Others have explored the possibility of using it with negative type
schemes~\cite{ahn11}.

@@

\bibliography{biblio}

>>>>

\subsection{Retractive-positive recursive types (\texttt{Mu.v})}
\label{sec:mu}

@@

\textbf{Termination.}  In some tools, like Coq, Agda, and Lean,
termination is checked statically, based on structural decrease.
Others, like Isabelle/HOL, allow one to write recursions first, and
prove (possibly with automated help) their termination
afterwards~\cite{krauss}.  Well-founded recursion replaces structural
decrease with decrease in a well-founded ordering. At least in
constructive type theory, evidence of well-foundedness then propagates
through code.  In contrast, our approach here, while less general,
does not clutter code with proofs.  For more on partiality and recursion
in theorem provers, see~\cite{bove+16}.

Subsidiary recursion can be seen as a generalization of \emph{nested
recursion}, which allows recursive calls of the form \verb|f (f x)|~\cite{krauss10}.
In subsidiary recursion, these are generalized to the form
\verb|f (g x)|, where \verb|g| could be \verb|f| or another
recursively defined function.

\textbf{Mendler encoding.}  Mendler introduced the basic idea of using
universal abstraction to support compositional termination checking;
an accessible source is~\cite{mendler91}.  He introduces a
functor-generic recursor of type $\all{X}{(\all{R}{(R \to X) \to F\ R
    \to X}) \to \mu\ F\ \to X}$. We have adopted this idea to the
constructor of the type \verb|Mu| (Section~\ref{sec:mu}).  Previous
works explored the categorical perspective on Mendler-style
recursion~\cite{uustalu99}, and its use with negative type
schemes~\cite{ahn11}.  Previous work from our group showed
how to derive inductive datatypes in Cedille using encodings
extending the Mendler encoding~\cite{firsov+18b,firsov+18a,stump18}.

@@

We have seen a derivation in Coq of a scheme for terminating
subsidiary recursion, where recursions may be nested and outer
recursive calls may be made on results of inner recursions.  We saw
examples invoking the \verb|span| function as a subsidiary recursion,
for functions \verb|wordsBy| and run-length encoding.  We also looked
briefly at the extension of this interface to support subsidiary
induction, with example lemmas about \verb|span|, and the decoding
correctness theorem for run-length encoding.  There are many other
interesting examples we can develop in Coq with this interface,
including natural-number division, which may invoke subtraction as a
subsidiary recursion.  Another example is Harper's regular-expression
matcher, which previous work showed can be implemented in Cedille
using a form of nested recursion that is subsumed by subsidiary
recursion~\cite{stump20}.  We may also attempt to extend the recursion
universe further, to allow other forms of recursion like
divide-and-conquer, where some (necessarily limited) ability to
recurse on values built using constructors is required.

\pagebreak
\bibliography{biblio}
